Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-02-03 draft_lm1.tex
expanded the discussion; added information to figure/table legends; see recent issues

Quantifying the attack ratio of disease is key to epidemiological inference and 
Public Health planning.
For multi-serotype pathogens, however, different levels of 
serotype-specific immunity make it difficult to assess the population at risk. 
In this paper we propose a Bayesian method for estimation of the attack ratio 
of an epidemic and the initial fraction of susceptibles using aggregated 
incidence data. 
We derive the probability distribution of the effective reproductive number, 
\rr, and use MCMC to obtain posterior 
distributions of the parameters of a single-strain SIR transmission model with 
time-varying force of infection.
Our method is showcased in a data set consisting of 18 years of dengue 
incidence in the city of Rio de Janeiro, Brazil.
We demonstrate that it is possible to learn about the initial fraction of 
susceptibles and the attack ratio even in the absence of serotype specific data.

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at  
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
Also, as the numerical integration routine implemented within BIP needs 
$\beta(t)$ to be available at 
arbitrary values of $t$, i.e., as continuous function of time because of the 
variable step size, we used linear interpolation to obtain values of $R_t$ for 
any time point.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubin's R (potential scale 
reduction), that approaches $1$ at convergence~\cite{brooks1998}.

@@

@@

Figure \ref{fig:rtseries}  shows the $R_t$ series, according to 
(\ref{eq:Rtestimate}). 
It can be seen that the inter-epidemic periods are characterized by $R_t$ being 
indistinguishable from 1. Due to the intrinsic variability of the \rr series, 
the examination of its credible intervals is essential to identify periods 
of sustained transmission. The wider intervals between epidemics are due to the 
scarcity of cases during these periods. Our methodology to derive the 
probability distribution of \rr is an original contribution applicable to the 
analysis of any incidence time-series of  any transmissible disease.

Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.
In the case of dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergence of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

The reason for the choice of a single strain model, is the lack of 
information regarding the actual proportions of each virus in 
circulation.
However, some information is available about the predominant 
serotypes for some epidemics in the period
of study~\cite{macedo_virological_2013}.

We sought to deal with all important sources of uncertainty impinging on the 
estimation of the AR of a dengue epidemic, but not all could be satisfactorily 
addressed in this analysis.
For instance, in any given epidemic there is a large number of 
mild and asymptomatic cases, which nevertheless acquire immunity.
It is estimated that for every case reported, up to 10-20 are not seen by health 
authorities (Oswaldo G. Cruz, personal communication).

Another source of uncertainty is under-reporting, which is a serious issue in
the health care systems of many developing countries such as Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of Dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, meaning that approximately 37\% of the 
suspected Dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways 
which are not easy to fully determine.

\begin{figure}
 \centering
 \includegraphics[width=16cm]{./plots/rt_series.png}

 \caption{{\bf Estimated time-series for $R_t$, along with 95\% credible 
intervals.} Top panel show reported cases from which $R_t$ is estimated.}
 \label{fig:rtseries}
\end{figure}

Figure \ref{Fig:S0}, shows the model from (\ref{eq:model}) fitted to the data. 
In it we can see that the susceptibles series in each epidemic starts at the 
estimated level of $S_0$.
The proportion of susceptibles may seem low, but we 
must remember that these estimates are being affected by an unknown 
under-reporting factor, which experts suggest is somewhere between 5 and 10, 
i.e. for every case observed there are 5 or 10 unobserved.
Since this under-reporting affects both the numerator and denominator of 
(\ref{eq:AR2}), its effects should cancel out, giving us an unbiased attack 
ratio 
estimate.
One other possible source of bias which would lead to the 
underestimation of $S_0$ could come from a significant part of the population 
not being exposed to the disease.
However, as we can see in figure \ref{fig:mapas} , the disease spreads pretty 
much homogeneously over 
the entire city (in the last four epidemics, at least).

\begin{figure}
 \subfigure[2010]{\includegraphics[width=8cm, 
height=4.82cm]{./plots/heatmap2010.jpg}
% heatmap2010.jpg: 1075x589 pixel, 96dpi, 28.44x15.58 cm, bb=0 0 806 442
}
  \subfigure[2011]{\includegraphics[width=8cm]{./plots/heatmap2011.jpg}}
  \subfigure[2012]{\includegraphics[width=8cm]{./plots/heatmap2012.jpg}
% heatmap2010.jpg: 1075x589 pixel, 96dpi, 28.44x15.58 cm, bb=0 0 806 442
}
  \subfigure[2013]{\includegraphics[width=8cm]{./plots/heatmap2013.jpg}}
  \caption{{\bf Maps showing the incidence of dengue in the city of Rio de 
Janeiro from 2010 to 2013.} Circles indicate individual notified cases. A 
heatmap is overlayed on the maps showing absolute density of cases.}
  \label{fig:mapas}
\end{figure}

Table \ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak. It is interesting to notice that the 
larger epidemics, in terms of peak size are not the one with the 
greater attack ratios. This stresses the importance of knowing the 
immunological structure of the population in order to truly assess the impact 
of any epidemic.

The authors hope that this evidence will motivate public health 
authorities to invest in annual serological surveys, to determine the 
susceptibility profile to each dengue virus as well as to estimate the 
under-reporting factor of the notification system.

\section*{Acknowledgements}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for \rr.
The authors are also grateful to Claudia T. 
Code\c{c}o for helpful discussions about the manuscript.

\section*{Appendix}
\subsection*{A remark on prior distributions and tail behaviour of the 
distribution of $R_t$}
\label{sec:tails}
There are a number of approaches to deriving the distribution of \rr.
Alternatively to the approach described in the main text~\cite{mantel}, one 
could use the conditional distribution of \rr on 
$Y_{t+1}$ and $Y_t$ as defined in equation A7 of Nishiura et al.\cite{nishiura}:
\begin{equation}
\label{seq:unorm}
f_{R}(R_{t}) = (Y_tR_{t})^{Y_{t+1}} e^{-Y_tR_{t}}
\end{equation}
Noticing the kernel of (\ref{seq:unorm}) is that of a gamma distribution with 
$a_2 = Y_{t+1}+1$ and $b_2 = Y_t$, we obtain a proper density from which to 
construct $c_{\alpha}(R_t)$, simply by computing the appropriate quantiles of 
said distribution.
 This density is
\begin{equation}
\label{seq:densityNishiura}
f_N(R_t| a_2, b_2) =  \frac{b_2^{a_2}}{\Gamma(a_2)} R_t^{a_2-1} e^{-b_2 R_t}
\end{equation}

@@

\bibliography{lm1}
\section*{Figure Legends}
\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{plots/concat_SI.png}
\end{center}
\caption{
{\bf Susceptibles and Infectious posterior curves.}  The curves were estimated 
only for the periods where $R_t> 1$.  The susceptible curves in the top panel 
reflect the prevalence of fraction of susceptibles to circulating strain(s) for 
each epidemic/outbreak. In the lower panel, we see the posterior distribution 
of infectious curves, represented by its Median and 95\% credible interval. The 
credible intervals are very narrow, and can be hard to distinguish from the 
median line. 
Scatter plot shows the observed cases. Prevalences (y-axes) are scaled as 
fractions of the entire population.
}
\label{Fig:S0}
\end{figure}

\section*{Tables}

\begin{table}[!ht]
\caption{
{\bf Median attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. Values are presented as percentage of total population. 
$^\dag$: Year corresponds to the start of the epidemic. 
Sometimes the peak of cases occur in the following year.
$^\ddag$: Susceptible fraction.}
\begin{center}
\begin{tabular}{c|c|c}
\hline
Year$^\dag$ & median Attack Ratio & $S_0^\ddag$ \\
\hline
1996 & 0.39 (0.17-0.54) & 0.00171(0.0012-0.0038)\\
1997 & 0.87 (0.74-0.87) & 0.00273(0.0027-0.0032)\\
1998 & 0.5 (0.49-0.5) & 0.00142(0.0014-0.0014)\\
1999 & 0.11 (0.037-0.2) & 0.00345(0.0018-0.01)\\
2000 & 0.25 (0.24-0.27) & 0.0155(0.015-0.016)\\
2001 & 0.48 (0.47-0.49) & 0.0495(0.048-0.051)\\
2005 & 0.15 (0.1-0.21) & 0.0147(0.01-0.021)\\
2006 & 0.11 (0.08-0.14) & 0.0281(0.022-0.037)\\
2007 & 0.15 (0.15-0.15) & 0.135(0.13-0.14)\\
2008 & 0.14 (0.031-0.31) & 0.00672(0.003-0.024)\\
2010 & 0.18 (0.17-0.19) & 0.0454(0.043-0.048)\\
2011 & 0.086 (0.082-0.094) & 0.215(0.2-0.23)\\
2012 & 0.14 (0.13-0.15) & 0.0621(0.058-0.068)\\

\hline
\end{tabular}

\end{center}

\label{tab:AR}
\end{table}

>>>>

Quantifying the attack ratio of disease is key to epidemiological inference and 
Public Health planning.
For multi-serotype pathogens, however, different levels of 
serotype-specific immunity make it difficult to assess the population at risk. 
In this paper we propose a Bayesian method for estimation of the attack ratio 
of an epidemic and the initial fraction of susceptibles using aggregated 
incidence data. 
We derive the probability distribution of the effective reproductive number, 
\rr, and use MCMC to obtain posterior 
distributions of the parameters of a single-strain SIR transmission model with 
time-varying force of infection.
Our method is showcased in a data set consisting of 18 years of dengue 
incidence in the city of Rio de Janeiro, Brazil.
We demonstrate that it is possible to learn about the initial fraction of 
susceptibles and the attack ratio even in the absence of serotype specific data.
On the other hand, the information provided by this approach is limited, 
stressing the need for detailed serological surveys to characterise the 
distribution of serotype-specific immunity in the population.

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at  
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
Also, as the numerical integration routine implemented within BIP needs 
$\beta(t)$ to be available at 
arbitrary values of $t$, i.e., as continuous function of time because of the 
variable step size, we used linear interpolation to obtain values of $R_t$ for 
any time point.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubin's R (potential scale 
reduction factor), which approaches $1$ at convergence~\cite{brooks1998}.

@@

Python and R code to perform all the analyses described above is publicly 
available at~\url{https://github.com/fccoelho/paperLM1}.

@@

In this paper we propose a method to bypass the lack of serotype-specific case 
data by informing the time-varying force of infection with the instantaneous 
reproductive number, \rr which we calculate from aggregated data.
Our contribution is two-fold: (i) we show a method to quantify uncertainty 
about \rr that is readily applicable to other diseases and; (ii) we propose to 
use \rr to inform a dynamic epidemic model with time-varying force of infection 
in order to gain insight into the attack ratio of each epidemic.

Figure~\ref{fig:rtseries}  shows the $R_t$ series, according to 
(\ref{eq:Rtestimate})~\cite{nishiura} along with the confidence bands derived 
in this paper. 
It can be seen that the inter-epidemic periods are characterized by $R_t$ being 
indistinguishable from 1.
Due to the intrinsic variability of the \rr series, the examination of its 
credible intervals is essential to identify periods of sustained transmission.
The wider intervals between epidemics are due to the scarcity of cases during 
these periods.
The method to quantify uncertainty proposed here provides more conservative 
credibility intervals, and therefore offers protection against false alarms.

\begin{center}
\begin{figure}[!h]
 \centering
 \includegraphics[width=16cm]{./plots/rt_series.png}
 \caption{{\bf Estimated time-series for $R_t$, along with 95\% credible 
intervals.} Top panel show reported cases from which $R_t$ is estimated.
As expected, uncertainty about \rr is greater when the case counts are low, for 
instance in the period $2003$--$2006$, which represented a big hiatus between 
major epidemics.
The intrinsic variability of \rr can be used to inform the time-varying force 
of infection, since it reflects variation in the vector population and other 
environmental factors such as temperature and seasonal variation.}
\label{fig:rtseries}
\end{figure}
\end{center}

A key epidemiological quantity is the attack ratio (AR) of an epidemic, a 
measure of morbidity and speed of spread which can be used to predict epidemic 
size and help efficient Public Health planning.
The AR depends fundamentally on the population at risk, which in the case of 
dengue is every naive (to a particular serotype) individual in the population.
Estimating the initial susceptible fraction $S_0$ for each epidemic is thus 
central to the estimation of the AR.
Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.
In the case of dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergence of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

Since there is very limited information regarding the actual proportions of 
each virus in circulation and most information available is about the 
predominant serotypes for some epidemics in the period of 
study only~\cite{macedo_virological_2013}, we propose the use of a 
simplified a single strain model.
The main argument we put forward is that by conditioning on the \rr series, we 
implicitly take into account the variability introduced by the 
co-circulation of multiple serotypes and heterogeneous levels of immunity in 
the general population.
We sought to deal with all important sources of uncertainty impinging on the 
estimation of the AR of a dengue epidemic, but not all could be satisfactorily 
addressed in this analysis.
For instance, in any given epidemic there is a large number of 
mild and asymptomatic cases, which nevertheless acquire immunity.
It is estimated that for every case reported, up to 10-20 are not seen by health 
authorities (Oswaldo G. Cruz, personal communication).
Another source of uncertainty is under-reporting of diagnosed cases, which is a 
serious issue in the health care systems of many developing countries such as 
Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of Dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, meaning that approximately 37\% of the 
suspected Dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways 
which are not easy to fully determine.

Figure~\ref{Fig:S0}, shows the model from (\ref{eq:model}) fitted to the data.
Despite its limitations, our simplified model fits the data well.
In it we can see that the susceptibles series in each epidemic starts at the 
estimated level of $S_0$.
The proportion of susceptibles may seem low, but we 
must remember that these estimates are being affected by an unknown 
under-reporting factor, which experts suggest is somewhere between 5 and 10, 
i.e. for every case observed there are 5 or 10 unobserved.
Since this under-reporting affects both the numerator and denominator of 
(\ref{eq:AR2}), its effects should cancel out, giving us an unbiased attack 
ratio estimate.
One other possible source of bias which would lead to the underestimation of 
$S_0$ could come from a significant part of the population not being exposed to 
the disease.
However, as we can see in Figure~\ref{fig:mapas}, despite the differences in 
intensity (incidence), the entire city seems to be at risk, with no 
particularly ``protected'' areas, at least in the last four epidemics.

\begin{center}
\begin{figure}
 \subfigure[2010]{\includegraphics[width=8cm, 
height=4.82cm]{./plots/heatmap2010.jpg}
% heatmap2010.jpg: 1075x589 pixel, 96dpi, 28.44x15.58 cm, bb=0 0 806 442
}
  \subfigure[2011]{\includegraphics[width=8cm]{./plots/heatmap2011.jpg}}
  \subfigure[2012]{\includegraphics[width=8cm]{./plots/heatmap2012.jpg}
% heatmap2010.jpg: 1075x589 pixel, 96dpi, 28.44x15.58 cm, bb=0 0 806 442
}
  \subfigure[2013]{\includegraphics[width=8cm]{./plots/heatmap2013.jpg}}
  \caption{{\bf Maps showing the incidence of dengue in the city of Rio de 
Janeiro from 2010 to 2013.} Circles indicate individual notified cases. A 
heatmap is overlayed on the maps showing absolute density of cases.
It can be seen that several areas of the city were affected and no region seems 
to be free of transmission risk.
This suggests that although transmission risk varies spatially, there is 
significant exposure over the entire city.}
  \label{fig:mapas}
\end{figure}
\end{center}

Table~\ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak.
It is interesting to notice that the larger epidemics, in terms of peak size are 
not the one with the greater attack ratios.
This stresses the importance of knowing the immunological structure of the 
population in order to truly assess the impact of any epidemic. [WHY?]

We hope that the results presented in paper will motivate public 
health authorities to invest in annual serological surveys, to determine the 
susceptibility profile to each dengue virus as well as to estimate the 
under-reporting factor of the notification system.

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{plots/concat_SI.png}
\end{center}
\caption{
{\bf Susceptibles and Infectious posterior curves.}  The curves were estimated 
only for the periods where $R_t> 1$.  The susceptible curves in the top panel 
reflect the prevalence of fraction of susceptibles to circulating strain(s) for 
each epidemic/outbreak.
In the lower panel, we see the posterior distribution 
of infectious curves, represented by its median and 95\% credible interval.
Credible intervals are very narrow, and can be hard to distinguish from the 
median line. 
Dots show the observed cases, scaled as fractions of the entire 
population.
}
\label{Fig:S0}
\end{figure}

\begin{table}[!ht]
\caption{
{\bf Median attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. 
Values are presented as percentage of total population. 
$^\dag$: Year corresponds to the start of the epidemic. 
Sometimes the peak of cases occur in the following year.
$^\ddag$: Susceptible fraction.}
\begin{center}
\begin{tabular}{c|c|c}
\hline
Year$^\dag$ & median Attack Ratio & $S_0^\ddag$ \\
\hline
1996 & 0.39 (0.17-0.54) & 0.00171(0.0012-0.0038)\\
1997 & 0.87 (0.74-0.87) & 0.00273(0.0027-0.0032)\\
1998 & 0.5 (0.49-0.5) & 0.00142(0.0014-0.0014)\\
1999 & 0.11 (0.037-0.2) & 0.00345(0.0018-0.01)\\
2000 & 0.25 (0.24-0.27) & 0.0155(0.015-0.016)\\
2001 & 0.48 (0.47-0.49) & 0.0495(0.048-0.051)\\
2005 & 0.15 (0.1-0.21) & 0.0147(0.01-0.021)\\
2006 & 0.11 (0.08-0.14) & 0.0281(0.022-0.037)\\
2007 & 0.15 (0.15-0.15) & 0.135(0.13-0.14)\\
2008 & 0.14 (0.031-0.31) & 0.00672(0.003-0.024)\\
2010 & 0.18 (0.17-0.19) & 0.0454(0.043-0.048)\\
2011 & 0.086 (0.082-0.094) & 0.215(0.2-0.23)\\
2012 & 0.14 (0.13-0.15) & 0.0621(0.058-0.068)\\
\hline
\end{tabular}
\end{center}
\label{tab:AR}
\end{table}

\newpage
\section*{Acknowledgements}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for \rr.
The authors are also grateful to Claudia T. 
Code\c{c}o for helpful discussions about the manuscript.

\newpage
\bibliography{lm1}

\section*{Appendix}
\subsection*{A remark on prior distributions and tail behaviour of the 
distribution of $R_t$}
\label{sec:tails}
There are a number of approaches to deriving the distribution of \rr.
Alternatively to the approach described in the main text~\cite{mantel}, one 
could use the conditional distribution of \rr on 
$Y_{t+1}$ and $Y_t$ as defined in equation A7 of Nishiura et 
al.~\cite{nishiura}:
\begin{equation}
\label{seq:unorm}
f_{R}(R_{t}) = (Y_tR_{t})^{Y_{t+1}} e^{-Y_tR_{t}}
\end{equation}
Noticing the kernel of (\ref{seq:unorm}) is that of a gamma distribution with 
$a_2 = Y_{t+1}+1$ and $b_2 = Y_t$, we obtain a proper density from which to 
construct $c_{\alpha}(R_t)$, simply by computing the appropriate quantiles of 
said distribution.
 This density is
\begin{equation}
\label{seq:densityNishiura}
f_N(R_t| a_2, b_2) =  \frac{b_2^{a_2}}{\Gamma(a_2)} R_t^{a_2-1} e^{-b_2 R_t}
\end{equation}

@@
